---
title: "Feature overview and tutorial"
output:
  html_document:
    df_print: paged
  rmarkdown::html_vignette:
    df_print: paged
  pdf_document:
    df_print: kable
  word_document: default
vignette: >
  %\\\\VignetteIndexEntry{Feature overview and tutorial}
  %\\\\VignetteEncoding{UTF-8}
  %\\\\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
knitr::opts_chunk$set(fig.width=6, fig.height=6)
```

## Setup

Load the package:

```{r}
library(taxplore)
```

### Snapshot configuration (optional)

The charts can be embedded in static documents (PDF, Word, etc.). See also [auto-snapshots](#auto-snapshots) below. Here we run the small small helper function `taxplore_configure_snapshot()`, which makes sure that snapshots are PNG images with a resolution of 300 DPI. This is only relevant if "knitting" this document to PDF or Word, etc. Otherwise, this has no effect.

```{r}
taxplore_configure_snapshot()
```

> **Note** This function modifies the `dev`, `dpi` and `screenshot.opts` settings for *all* chunks. However, it is always possible to override the settings for individual chunks.

## Data formats

### Long format without abundances

The example dataset used in this tutorial contains 54 records of macrofungi identified during two surveys on a grassland plot in Switzerland.

The `grasslandfungi.records` version of the dataset simply consists of taxonomic lineages, one row per record (taxa can be repeated):

```{r}
data(grasslandfungi.records)
head(grasslandfungi.records)
```

The taxonomic hierarchy of the records from the two visits can easily be displayed, whereby the widths are proportional to taxa abundances:

```{r cache=F}
plot_krona(grasslandfungi.records)
```

### Hierarchy + abundances

Another version of the dataset (`grasslandfungi`) has aggregated taxonomic lineages along with taxa abundances for both visits, as well as the total record count:

```{r}
data(grasslandfungi)
head(grasslandfungi[c('family', 'species', 'visit_1', 'visit_2', 'n_total')])
```

If the taxa are simply provided without abundances, every one of them obtains the same weight in the chart as well (=visualization of presences/detections):

```{r width=4, height=5}
tax_ranks <- names(grasslandfungi.records)
plot_krona(grasslandfungi[tax_ranks])
```

To scale by abundance, we supply `n_total` as second argument:

```{r width=4, height=5}
plot_krona(grasslandfungi[tax_ranks], grasslandfungi$n_total)
```

> The second `magnitude` argument can also be a *matrix/data frame* with abundances from different surveys (samples, datasets), which can optionally be displayed separately (see [datasets](#datasets), more details in the `make_krona()` documentation). By default, taxa widths are scaled by the row totals. Consequently, supplying `grasslandfungi[c('visit_1', 'visit_2')]` as `magnitude` gives the same result as supplying `grasslandfungi$n_total`.

### Phyloseq objects

The popular [phyloseq](https://joey711.github.io/phyloseq) format is supported out of the box, see `vignette("phyloseq")`.

### Hierarchical list data structure

Alternatively, the classification can be a hierarchical data structure composed of nested lists:

```{r width=5, height=5}
classification <- list(
  name = 'Root',
  n = 15,
  list(
    name = 'A',
    n = 10,
    list(name = 'A.1', n = 5),
    list(name = 'A.2', n = 3)
  ),
  list(name = 'B', n = 5)
)
plot_krona(classification)
```

> The node attribute `name` is mandatory. The `n` attribute indicates the magnitude and can be omitted or named differently (`magnitude = 'othername'`)

### Pre-generated HTML charts

Finally, it is possible to directly display file containing a Krona chart:

```{r}
plot_krona(file = 'https://krona.sourceforge.net/examples/xml.krona.html')
```

... or `plot_krona()` can display HTML code stored in a character string. In this example, the HTML is generated by `make_krona()`:

```{r}
chart_html <- make_krona(classification)
plot_krona(chart_html)
```

## Chart generation options

So far, `plot_krona()` has been used to directly display the Krona charts as [HTML widget](http://htmlwidgets.org). Behind the scenes, `make_krona()` acts as the "working horse" for generating the chart code. This function can be used in the same way as `plot_krona()`, but returns the HTML as character string instead of a HTML widget. Both functions can also save the chart to a HTML file with `outfile`:

```{r eval = FALSE}
# save HTML in character string
chart_html <- make_krona(classification)
# display AND save the chart
plot_krona(chart_html, outfile = 'chart.html')
# save the chart
make_krona(classification, outfile = 'chart.html')
```

The two functions offer almost all features of the original [Krona software](https://github.com/marbl/Krona/wiki), as described in the following. The generated charts are identical or almost identical as the ones generated by Krona.

### Datasets

Multiple datasets can grouped within the same chart, selectable from a list in the **top-left corner**.

In this example the two separate surveys from the grassland fungi dataset are visualized separately (`visit_1` and `visit_2`). To prevent the aggregation of the counts across the `visit_1` and `visit_2` magnitude columns, we have to specify `dataset_group = 'separate'`:

```{r}
plot_krona(grasslandfungi[tax_ranks],
       grasslandfungi[c('visit_1', 'visit_2')],
       dataset_group = 'separate')
```

> In static contexts without user interaction (PDF, Word, etc.), a snapshot of the first dataset (*visit_1*) is taken.

> `dataset_group` can also be a character vector of length `ncol(magnitude)`; for phyloseq objects `group_vars` can be specified (see vignette("phyloseq")`).

### Data-dependent custom coloring

See `vignette("phyloseq")`

### Attributes: Displaying extra information

Additional taxon-specific information can be displayed in the **top-right corner** when clicking on a node of the hierarchy. This extra-information is configured through `ChartAttribute()`. 

For example, the grassland fungi dataset has three more columns with useful information: the current IUCN redlist status in Switzerland, and taxon IDs in the [GBIF](https://www.gbif.org) and [SwissFungi](https://www.wsl.ch/map_fungi) databases: 

```{r}
head(grasslandfungi[c('species', 'redlist_CH', 'gbif_taxon_id', 'swissfungi_taxon_id')])
```

We can now add all the redlist status and GBIF + SwissFungi links to the chart (explore by clicking on taxa and observing the top-right corner):

```{r}
attrs <- list(
  ChartAttribute('Redlist status', data = grasslandfungi$redlist_CH),
  ChartAttribute(
    'GBIF',
    data = as.character(grasslandfungi$gbif_taxon_id),
    hrefBase = 'https://www.gbif.org/species/',
    missing_value = NA_character_
  ),
  ChartAttribute(
    'SwissFungi',
    data = as.character(grasslandfungi$swissfungi_taxon_id),
    # to display text instead of the ID (be aware that this makes HTML files larger)
    # text = 'SwissFungi',
    hrefBase = 'https://www.wsl.ch/map_fungi/search?taxon='
  )
)

plot_krona(grasslandfungi[tax_ranks], grasslandfungi$n_total, attributes=attrs,
       total_label = 'Records')
```

> `ChartAttribute()` offers some more options for customizing URLs and displayed text

> **Note on higher taxonomic ranks**: If numeric IDs were provided, they would be averaged at higher ranks, which produces incorrect IDs for the *genus*, *family*, etc. ranks. Therefore, we specified `as.character(taxon_id)`; as text data is only shown if all child nodes are consistent. Alternatively, specify `only_ranks = 'species'`. See also *summary_fn* in `ChartAttribute()`.

Even adding extra information for only one or a few nodes in the hierarchy is possible, assuming that both the classification *and* the `magnitudes` data have row names to identify them. Here we link the global redlist page for [*Ramaria broomei = Phaeoclavulina macrospora*](https://redlist.info/iucn/species_view/111880) (zoom into Gomphaceae and click on the species to see the link):

```{r eval = FALSE}
attrs <- list(
  #...
  ChartAttribute(
    'Global redlist',
    data = c(Phmc = '111880'),
    text = 'VU C2a(i)',
    hrefBase = 'https://redlist.info/iucn/species_view/'
  )
)

plot_krona(grasslandfungi[tax_ranks],
       grasslandfungi[,'n_total', drop = FALSE],
       attributes = attrs)
```

#### Attributes in hierarchical list classification

In a [hierarchical list data structure](#hierarchical-list-data-structure), attribute information is easily added:

```{r eval = FALSE}
classification <- list(
  name = 'Root',
  n = 15,
  list(
    name = 'A',
    n = 10,
    info = 'Info about A'
  )
  #(...)
)
plot_krona(classification, attributes = list(
  ChartAttribute('info', displayName = 'Information')  # displayName is optional
))
```


### Advanced HTML chart options

Behind the scenes, `make_krona()` assembles an [XML document](https://github.com/marbl/Krona/wiki/Krona-2.0-XML-Specification) document, which is then passed to `generate_krona_html()`.

#### Use KronaTools instead

To use the official [Krona toolset](https://github.com/marbl/Krona) instead of the internal method, specify  `make_krona(..., method = 'kronatools')` (see `kt_import_xml()`). The result should usually be the same.

#### Remote resources instead of self-contained HTML

By default, Krona charts are self-contained HTML files that can be displayed without internet connection. Images and the JavaScript code are embedded in every file. If generating many files, an external resources directory can be specified (see `KronaOpts()`):

```{r eval = FALSE}
plot_krona(grasslandfungi.records, resources_url = 'https://marbl.github.io/Krona/')
```

> Use `set_krona_opts()` to set this option for all following charts.

To copy the assets delivered with this R package (= latest Krona code and some images, [`d1479b3`](https://github.com/marbl/Krona/commits/master)) to a target directory, call:

```{r eval = FALSE}
copy_krona_resources('resources')
```

#### Optimize HTML file size

A little space can be saved by including a "minified" version of the JavaScript code:

```{r eval = FALSE}
plot_krona(classification, minify = TRUE)
set_krona_opts(minify = TRUE)
```

## Styling embedded charts

The appearance of Krona charts can be changed with buttons and checkboxes, etc., and default values can be configured for every of these. If displaying the chart in a browser, the initial appearance can be configured with URL parameters added to the URL (`...?key1=value1&key2=value2`), e.g. http://marbl.github.io/Krona/examples/phymmbl.krona.html?depth=3&font=15.

For charts embedded in R-Markdown, `display = ChartDisplayOpts(...)` can be specified:

```{r cache = FALSE}
plot_krona(grasslandfungi.records,
       display = ChartDisplayOpts(font = 15, showMagnitude = TRUE))
```

> In R-Markdown chunks, set `cache = FALSE` if adjusting display options has no effect.

To make all following charts look the same, use:

```{r eval = FALSE}
set_chart_display_opts(font = 15, showMagnitude = TRUE)
```


## Auto-snapshots

In "static" documents such as PDF, Word, etc., snapshots are automatically taken. In this "snapshot mode", no buttons and other controls are shown. This behavior can be forced (or prevented) with the `interactive` option in `plot_krona()`:

```{r cache = FALSE}
plot_krona(grasslandfungi.records, interactive = FALSE)
```

It is also possible to force a snapshot in interactive contexts by setting the R-Markdown chunk option `screenshot.force = TRUE` *and* calling `plot_krona()` with `interactive = FALSE`. To do this globally for all chunks:

```{r eval = FALSE}
set_chart_opts(interactive = FALSE)
knitr::opts_chunk$set(screenshot.force = TRUE)
```

To create a snapshot image (not HTML file) manually:

```{r eval = FALSE}
make_krona(grasslandfungi.records, 
           outfile = 'chart.html',
           snapshot = TRUE)
webshot2::webshot('chart.html', 'chart-snapshot.png', 672, 480)
# file.remove('chart.html')
```

### Details on snapshot formats

On top of this tutorial, the snapshot format was set to PNG. In PDF documents, the default would be to embed the charts as vector graphics and text. But generating PNG images instead can be useful due to different reasons:

- file sizes can be smaller for complex charts
- thin lines may look nicer
- the resolution might not be correct (see  [issue](https://github.com/yihui/knitr/issues/2276#issuecomment-1686298985))

But you can still try without these settings (leave `dev = NULL`) to see how it looks like. Also, the PNG dimensions might not be correct (see also [below](#auto-snapshots)).

### Plot dimensions

The charts may be sized differently depending on the context:

**Inline view in R-Markdown chunks (RStudio)**: Always the same size

**HTML documents**: Charts are sized with `fig.width` and `fig.height` chunk options (in inches)

**PDF/Word documents**: HTML widget snapshots seem to always be scaled to the **full width** in the output, but still the image resolution and font size are determined by  `fig.width`/`fig.height` and `dpi` settings. Figure dimensions around 6-7 inches should look as expected (consider setting this at the top of the document: `knitr::opts_chunk$set(fig.width=6, fig.height=6)`). To make plots smaller, it is possible to modify `out.width` (e.g. `knitr::opts_chunk$set(out.width = '50%')`). However, the resolution (`dpi`) and font size (`ChartDisplayOpts(font = 15)`) may need to be adjusted.

## Shiny apps

This example code shows how to use these charts in a Shiny app (opens in a separate window). It also requires *phyloseq* to be installed.

```{r eval = FALSE}
library(shiny)
library(phyloseq)
library(taxplore)

data(GlobalPatterns, enterotype)

# each dataset is a list(taxonomy, magnitude)
datasets <- list(
  `Survey of grassland fungi` = list(grasslandfungi[tax_ranks], grasslandfungi$n_total),
  `Global patterns of 16S diversity` = list(tax_table(GlobalPatterns), taxa_sums(GlobalPatterns)),
  `Enterotypes of the human gut microbiome` = list(tax_table(enterotype), taxa_sums(enterotype))
)

ui <- bslib::page_fluid(
  titlePanel('Visualize community datasets'),
  radioButtons(
    'dataset',
    'Select dataset',
    choices = names(datasets),
    selected = names(datasets)[1]
  ),
  checkboxInput('use_abund', 'Use abundances', value = TRUE),
  KronaChartOutput('chart', height='40em')
)

server <- function(input, output) {
  output$chart <- renderKronaChart({
    d <- datasets[[input$dataset]]
    plot_krona(d[[1]], magnitude = if (input$use_abund) d[[2]])
  })
}

shinyApp(ui = ui, server = server)
```



